---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  fig.height=5, fig.width=8, 
  message=FALSE, warning=FALSE,
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# sfo

```{r, include=FALSE}
library(plotly)
library(dplyr)
library(sfo)
p <- sfo_passengers %>%
  group_by(activity_period) %>%
  summarise(total = sum(passenger_count), .groups = "drop") %>%
  mutate(date = as.Date(paste(substr(activity_period, 1, 4), 
                              substr(activity_period, 5, 6), "01", sep = "/"))) %>%
  plot_ly(x = ~ date, y = ~ total,
          type = "scatter", 
          mode = "lines") %>%
  layout(title = "Monthly Air Traffic Passengers in SFO",
         yaxis = list(title = "Number of Passengers"),
         xaxis = list(title = "Source: San Francisco data portal (DataSF)"))

orca(p, "man/figures/total.svg")  
```


<!-- badges: start -->
[![CRAN\_Status\_Badge](https://www.r-pkg.org/badges/version/sfo)](https://cran.r-project.org/package=sfo)
[![lifecycle](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![License: MIT](https://img.shields.io/badge/License-MIT-blue.svg)](https://opensource.org/licenses/MIT)
[![GitHub commit](https://img.shields.io/github/last-commit/RamiKrispin/sfo)](https://github.com/RamiKrispin/sfo/commit/master)
<!-- badges: end -->



The **sfo** package provides summary of the monthly passengers in San Francisco International Airport (SFO).

**Note:** WIP...

<img src="man/figures/total.svg" width="90%" />
## Installation

WIP still not available in CRAN but can be install from Github:

``` r
# install.packages("devtools")
devtools::install_github("RamiKrispin/sfo")
```
## Usage

The `sfo_passengers` dataset provides a monthly summary of the number of passengers in SFO airport by different categories (such as terminal, geo, type, etc.):

```{r example}
library(sfo)

data("sfo_passengers")

head(sfo_passengers)
```

Below are some use cases of the package.

### Total number of passngers

The total number of passengers in most recent month by `activity_type_code` and `geo_region`:

```{r }
library(dplyr)

sfo_passengers %>%
  filter(activity_period == max(activity_period)) %>%
  group_by(activity_type_code, geo_region) %>%
  summarise(total = sum(passenger_count), .groups = "drop")
```

Likewise, we can summarize the total number of passengers in most recent month by airline:

``` r
library(plotly)

sfo_passengers %>%
  filter(activity_period == max(activity_period)) %>%
  group_by(published_airline) %>%
  summarise(total = sum(passenger_count), .groups = "drop") %>%
  arrange(-total) %>%
  mutate(published_airline = factor(published_airline,levels = published_airline)) %>%
  plot_ly(x = ~ published_airline, y = ~ total, type = "bar") %>%
  layout(title = "Monthly Air Traffic Passengers in SFO By Airline",
         yaxis = list(title = "Number of Passengers"),
         xaxis = list(title = ""))

```


```{r, include=FALSE}
library(plotly)
p <- sfo_passengers %>%
  filter(activity_period == max(activity_period)) %>%
  group_by(published_airline) %>%
  summarise(total = sum(passenger_count), .groups = "drop") %>%
  arrange(-total) %>%
  mutate(published_airline = factor(published_airline,levels = published_airline)) %>%
  plot_ly(x = ~ published_airline, y = ~ total, type = "bar") %>%
  layout(title = "Monthly Air Traffic Passengers in SFO By Airline",
         yaxis = list(title = "Number of Passengers"),
         xaxis = list(title = ""))
 orca(p, "man/figures/by_airline.svg") 
```

<img src="man/figures/by_airline.svg" width="100%" />
