---
title: "Introduction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{intro}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The **sfo** package provides a monthly summary of San Francisco International (SFO) Airport air traffic passengers.

The `sfo_passengers` dataset contains the following fields:

-   `activity_period` - Activity year and month in YYYYMM format
-   `operating_airline` -Airline name for the operator of aircraft
-   `operating_airline_iata_code` -The International Air Transport Association (IATA) two-letter designation for the Operating Airline
-   `published_airline` -Airline name that issues the ticket and books revenue for passenger activity
-   `published_airline_iata_code` -The International Air Transport Association (IATA) two-letter designation for the Published Airline
-   `geo_summary` -Designates whether the passenger activity in relation to SFO arrived from or departed to a location within the United States ("domestic"), or outside the United States ("international") without stops
-   `geo_region` -Provides a more detailed breakdown of the GEO Summary field to designate the region in the world where activity in relation to SFO arrived from or departed to without stops
-   `activity_type_code` -A description of the physical action a passenger took in relation to a flight, which includes boarding a flight ("enplanements"), getting off a flight ("deplanements") and transiting to another location ("intransit")
-   `price_category_code` -A categorization of whether a Published Airline is a low-cost carrier or not a low-cost carrier}
-   `terminal` -The airport terminal designations at SFO where passenger activity took place
-   `boarding_area` -The airport boarding area designations at SFO where passenger activity took place
-   `passenger_count` -The number of monthly passengers associated with the above attribute fields

Data source: San Francisco data portal (DataSF) [API](https://data.sfgov.org/Transportation/Air-Traffic-Passenger-Statistics/rkru-6vcg)

```{r}
library(sfo)

data("sfo_passengers")

str(sfo_passengers)

head(sfo_passengers)

```
