---
title: "Analyzing SFO Passengers"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{analyzing_sfo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  fig.height=5, fig.width=8, 
  message=FALSE, warning=FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

The `sfo_passengers` dataset provides monthly statistics about the air traffic passengers at San Francisco International Airport (SFO) between July 2005 and December 2022. This vignette focuses on exploratory analysis approaches for this dataset.

### Data prep

We will start with reviewing the data structure:

```{r setup}
library(sfo)

data("sfo_passengers")
```

For simplicity, let's use a shorter name, `d`, for the dataset:

```{r}

d <- sfo_passengers

```

Next let's review the characteristics of the dataset:

```{r }

head(d)

str(d)

```

The `activity_period` variable provides the year-month of the observation using integer format. That is convenient if we filter by year and month, but less convenient if we want to plot the data as time series. Therefore, we will reformat the variable from YYYYMM to YYYYMMDD, setting the month's day as 1st with the `as.Date` function:

```{r}
d$date <- as.Date(paste(substr(d$activity_period, 1,4), substr(d$activity_period, 5,6), "01", sep ="/"))

head(d[,c("activity_period", "date")])

min(d$date)
max(d$date)

```

### Exploratory analysis

After we reviewed and reformatted the data, we can start exploring the data and answer some questions such as:

-   The monthly number of passengers at SFO

-   The airlines' distribution

-   The geographic distribution of the passengers

#### Monthly number of passengers

There are multiple ways to aggregate the datasets for monthly time-series data, such as the passengers' travel type (arrival vs. departure), geo type, and their combination. Let's start with the total aggregation of the data:


```{r}
library(dplyr)
library(plotly)

d_total <- d %>%
  group_by(date) %>%
  summarise(total = sum(passenger_count))

head(d_total)

plot_ly(data = d_total,
        x = ~ date,
        y = ~ total,
        type = "scatter",
        mode = "lines") %>%
  layout(title = "Monthly Air Traffic Passengers at SFO",
         yaxis = list(title = "Number of Passengers"),
         xaxis = list(title = "Source: San Francisco data portal (DataSF)"))

```

The COVID-19 pandemic effect can be seen in the series above, as the number of passengers sharply dropped in March 2020.

Likewise, we can aggregate the data by the activity type (arrival, departure, transit):

```{r}

d_activity_type <- d %>%
  group_by(date, activity_type_code) %>%
  summarise(total = sum(passenger_count), .groups = "drop") %>%
  as.data.frame()

head(d_activity_type)

plot_ly(data = d_activity_type,
        x = ~ date,
        y = ~ total,
        color = ~ activity_type_code,
        type = "scatter",
        mode = "lines") %>%
  layout(title = "Monthly Air Traffic Passengers at SFO by Activity Type",
         yaxis = list(title = "Number of Passengers"),
         xaxis = list(title = "Source: San Francisco data portal (DataSF)"))

```

The following plot provides the distribution of passengers by the flight geo type (international vs. domestic). We will use the `pivot_wider` function from the **tidyr** package to transform the `data.frame` into a wide format:

```{r}
library(tidyr)

d_geo <- d %>%
  group_by(date, geo_summary) %>%
  summarise(total = sum(passenger_count), .groups = "drop") %>%
  pivot_wider(names_from = geo_summary, values_from = total) %>%
  as.data.frame()

head(d_geo)

plot_ly(data = d_geo, 
        x = ~ date, 
        y = ~ Domestic, 
        name = "Domestic", 
        type = "scatter", 
        mode = "none", 
        stackgroup = "one", 
        groupnorm = "percent", fillcolor = "#4C74C9") %>%
  add_trace(y = ~ International,
            name = "International",
            fillcolor = "#50CB86") %>%
  layout(title = 'United States Personal Expenditures by Categories',
         yaxis = list(title = "Proportion of Passengers", 
                      showgrid = FALSE,
                      ticksuffix = "%"),
         xaxis = list(title = "Source: San Francisco data portal (DataSF)",
                      showgrid = FALSE))


```

As expected, the distribution of passengers by geo type has changed since the start of the COVID-19 pandemic, increasing domestic flights' proportion.

#### Airlines Analysis

The dataset provides two indicators about the airline company:

-   `operating_airline` field describes the aircraft operator's name, and

-   `published_airline` filed describes the name of the airline that issues the ticket and books revenue

In most cases, it is the same company. In other cases, such as code-sharing operators, it could be two different companies. Let's review the characteristics of those variables:

```{r}

length(unique(d$operating_airline))
length(unique(d$published_airline))
```

You can notice that there are fewer published airlines compared to operating airlines. This potentially means that some operating airlines have lines with code sharing. For simplicity, we will focus on the `operating_airline` variable. Let's review the unique combination of this variable:

```{r}
sort(unique(d$operating_airline))

```

The main thing that immediately pops out of this list is that United Airlines, the main carrier at SFO, is listed twice as `United Airlines` and `United Airlines - Pre 07/01/2013`. I could not find any reference that explains the second label. If analyzing data before July 2013, I would relabel those cases as `United Airlines`. For this analysis, we will focus on 2022, the most recent full year. Let's start by filtering the data for 2022 flights:

```{r}

d22 <- d %>% filter(activity_period >= 202201 & activity_period < 202301) %>%
  group_by(published_airline, geo_summary) %>%
  summarise(total = sum(passenger_count), .groups = "drop")

```

Before we summarize the data by airlines, let's see the distribution of flights by geo type (domestic vs. international):

```{r}

d22 %>% group_by(geo_summary) %>%
  summarise(total_geo = sum(total)) %>%
  plot_ly(labels = ~ geo_summary,
          values = ~ total_geo) %>%
  add_pie(hole = 0.6)
```

More than 75% of the air passenger traffic in SFO during 2022 were domestic. That is close to 47M passengers out of 60.8M passengers. Let’s use a treemap plot to see the distribution of passengers by airline and geo:

```{r}
  plot_ly(data = d22 %>% filter(geo_summary == "Domestic"),
          labels = ~ published_airline, 
          values = ~ total,
          type = "treemap",
          parents=  ~geo_summary,
          domain = list(row=0),
          name = "Domestic",
          textinfo="label+value+percent parent") %>%
    add_trace(data = d22 %>% filter(geo_summary == "International"),
            labels = ~ published_airline, 
            values = ~ total,
            type = "treemap",
            parents=  ~geo_summary,
            domain = list(row=1),
            name = "International",
            textinfo="label+value+percent parent") %>%
    layout(title = "Passengers Distribution During 2022 by Geo",
      grid=list(columns=1, rows=2))
```

### Air Traffic Summary

Last but not least, it would be interesting to leverage the different categorical variables to create a summary plot with Sankey. We will use the `sankey_ly` function, a **plotly** wrapper that automatically creates the data transformation and a Sankey plot with **plotly**. We will rank the top 20 airlines during 2022 and plot the passengers' distribution by the airline, travel type (domestic or international), travel geo, activity type (deplaned, enplaned, and transit), and fare type (low or other):

```{r}
rank <- sfo_passengers %>%
  filter(activity_period >= 202201 & activity_period <= 202212) %>%
  group_by(operating_airline) %>%
  summarise(total = sum(passenger_count), .groups = "drop") %>%
  arrange(-total) %>%
  slice_head(n = 20) %>%
  select(operating_airline)


sfo_passengers %>%
  filter(activity_period >= 202201 & activity_period <= 202212, operating_airline %in% rank$operating_airline) %>%
  group_by(operating_airline, geo_summary, geo_region, activity_type_code, 
           price_category_code, terminal) %>%
  summarise(total = sum(passenger_count), .groups = "drop") %>%
  sankey_ly(cat_cols = c("operating_airline","geo_summary",
                      "geo_region", "activity_type_code",
                      "price_category_code"),
          num_col = "total",
          title = "SFO Air Traffic Passengers Dist. During 2022")
```

